Coupling finite and boundary element methods to solve the Poisson-Boltzmann equation for electrostatics in molecular solvation

### 2022-03-15 methods.tex
Updated plots and manuscript
I attach plots with preconditioner and without. It give a little more understanding of behaviour of each method, at least I think so. We should discuss that.
In terms of manuscript, I am sorry, it is quite rough draft, but I am struggling with what should be included in terms of mathematics and derivation of methods.

The implicit solvent model can be described mathematically as a coupled system of partial differential equations, where the Poisson-Boltzmann governs in the solvent region ($\Omega_X$ in Figure XX), and the Poisson equation in the solute region ($\Omega_Y$ in Figure XX). These regions are interfaced by the molecular surface ($\Gamma$), where the potential ($\phi$) and electric displacement ($\epsilon\partial\phi/\partial\mathbf{n}$) are continuous. 
%
\begin{align}\label{eq:pbe}
\nabla^2\phi_1(\mathbf{x}) &= \frac{1}{\epsilon}\sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2 &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1 = \phi_2 &\quad \epsilon_1\frac{\partial\phi_1}{\partial\mathbf{n}} = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}} \quad \mathbf{x}\in \Gamma. 
\end{align}
%
where $\epsilon_1$ and $\epsilon_2$ are the dielectric constants in the solute and solvent, respectively, $\kappa$ is the inverse of the Debye length, related with the salt concentration, and $q_k$ are the values of the partial charges, located at $\mathbf{x}_k$.

The electrostatic potential in $\Omega_1$ can be further decomposed into singular and regular components as $\phi_1 = \phi_c + \phi_R$, where $\phi_c$ is the solution to
%
\begin{align}\label{eq:phic}
\nabla^2\phi_c(\mathbf{x}) &= \frac{1}{\epsilon_1}\sum_{k=1}^{N_q}q_k\delta(\mathbf{x},\mathbf{x}_k) \quad \mathbf{x}\in\Omega_1\cup\Omega_2\nonumber\\
\phi_c(\mathbf{x})&=0 \quad \text{ as } |\mathbf{x}|\to\infty
\end{align}

@@

\begin{itemize}
    \item Description with constant and variable permittivity 
    \item Standard BEM-BEM and FEM-BEM coupling approach
    \item Hybrid FEM-BEM coupling
\end{itemize}

>>>>

The implicit solvent model can be described mathematically as a coupled system of partial differential equations, where the Poisson-Boltzmann governs in the solvent region ($\Omega_X$ in Figure XX), and the Poisson equation in the solute region ($\Omega_Y$ in Figure XX). These regions are interfaced by the molecular surface ($\Gamma$), where the potential ($\phi$) and electric displacement ($\epsilon\partial\phi/\partial\mathbf{n}$) are continuous. 
%
\begin{align}\label{eq:pbe}
\nabla^2\phi_1(\mathbf{x}) &= \frac{1}{\epsilon_1}\sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2 (\mathbf{x}), &\quad \epsilon_1\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
%
where $\epsilon_1$ and $\epsilon_2$ are the dielectric constants in the solute and solvent, respectively, $\kappa$ is the inverse of the Debye length, related with the salt concentration, and $q_k$ are the values of the partial charges, located at $\mathbf{x}_k$.

The electrostatic potential in $\Omega_1$ can be further decomposed into singular and regular components as $\phi_1 = \phi_c + \phi_R$, where $\phi_c$ is the solution to
%
\begin{align}\label{eq:phic}
\nabla^2\phi_c(\mathbf{x}) &= \frac{1}{\epsilon}\sum_{k=1}^{N_q}q_k\delta(\mathbf{x},\mathbf{x}_k) \quad \mathbf{x}\in\Omega_1\cup\Omega_2\nonumber\\
\phi_c(\mathbf{x})&=0 \quad \text{ as } |\mathbf{x}|\to\infty
\end{align}

@@

\begin{itemize}
    \item Description with constant and variable permittivity

BEM reduces the dimension of the problem by using the boundary integral equation, hence its popularity as a numerical solution of  the Poisson-Boltzmann equation in molecular electrostatics. Unfortunately, such advantage comes also with the cost. A fundamental solution must be found before the BEM can be applied. There are many linear problems for which fundamental solutions are not known. It is a case if we consider the heterogenous permittivity inside the molecule
   \begin{align} \label{eq:pbe_vp}
\epsilon_1(\mathbf{x}) \nabla^2\phi_1(\mathbf{x}) &= \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k) \quad  \mathbf{x} \in \Omega_1\nonumber\\
\left(\nabla^2 - \kappa^2\right)\phi_2(\mathbf{x})  &= 0 \quad\mathbf{x}\in\Omega_2\nonumber\\
\phi_1(\mathbf{x})  = \phi_2(\mathbf{x}),  &\quad \epsilon_1(\mathbf{x})\frac{\partial\phi_1}{\partial\mathbf{n}}(\mathbf{x})  = \epsilon_2\frac{\partial\phi_2}{\partial\mathbf{n}}(\mathbf{x})  \quad \mathbf{x}\in \Gamma. 
\end{align}
For such case finite element methods (FEM) are more suitable.

The coupling of finite element (FEM) and boundary element (BEM) methods is the most widely used approach for solving multi-physical problems on an unbounded domain. It allows to take advantage of both methods. On the one hand, the BEM approximates only boundary conditions, hence it is commonly used for problems involving infinite or semi-infinite domains. On the other hand, the FEM is known for its robustness and universal applicability even for problem of inhomogeneous or non-linear nature.
And that is why, we present two formualtions of FEM-BEM coupling for Poisson-Boltzmann equation.

    \item Standard FEM-BEM coupling approach

    We start with the variational formulation of the internal problem. Applying integration by parts for first equation of~\eqref{eq:pbe_vp} for every $v \in H_0^1(\Omega_1)$ we have
\begin{equation}
\label{eq:fem}
 \int_{\Omega_1} \epsilon_1 \nabla \phi_1 : \nabla v ~d\mathbf{x}  - \oint_\Gamma \epsilon_1\partial_n \phi_1 v ds =   \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)  v ~d\mathbf{x}.
\end{equation}
For external problem we use BEM and the same direct formulation as in~\eqref{eq:direct}, hence we obtain
\begin{align*}
\frac{\phi_{1,\Gamma}}{2} - K_{Y}^{\Gamma}(\phi_{1,\Gamma}) + \frac{\epsilon_1}{\epsilon_2}V_{Y}^{\Gamma} \left( \frac{\partial}{\partial \mathbf{n}} \phi_{1,\Gamma} \right) & = 0.
\end{align*}

    \item Hybrid FEM-BEM coupling

    The main difference between the standard and hybrid FEM-BEM formulations is addition of hybrid variable $\widetilde{\phi}$ that is a trace of $\phi_1$ and $\phi_2$ on the interface $\Gamma$ ($\widetilde{\phi} = \phi_1 = \phi_2$ on  $\Gamma$). Both FEM and BEM formulations are modified to include the hybrid variable by adding some terms that enforce the equality of it with the trace of $\phi_1$ and $\phi_2$.
    The previous standard formulation of interior problem~\eqref{eq:fem} can be rewritten as follows
\begin{align*}
 \int_{\Omega_m} \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \nabla \phi_1 : \nabla v d\mathbf{x} -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2}  \partial_n \phi_1 v ds
  & \\
 -  \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} \partial_n v (\phi_1 - \widetilde{\phi}) ds
  + \tfrac{\tau_F}{h} \oint_\Gamma \frac{\epsilon_1(\mathbf{x})}{\epsilon_2} (\phi_1 - \widetilde{\phi}) v ds & = \int_{\Omega_1}  \sum_{k=1}^{N_q} q_k\delta(\mathbf{x},\mathbf{x}_k)   v ~d\mathbf{x}.
\end{align*}
With every iteration we are solving above problem for given $\widetilde{\phi}$ .
%We are using FEniCS finite element library, hence the above formulation is more sumilar to the implementation.

In the case of exterior problem of~\eqref{eq:pbe_vp} we use Yukawa kernel and 
%matrix form, for clarity and similarity with BEMpp implementation. In this case we consider linearised formulation of the Poisson-Boltzman equation that means our operators are built as for modified Helmholtz equation with wavenumber $\kappa$.
For given $\widetilde{\phi}$, with each iteration we are solving the following system
\begin{align*}
V_{Y}^{\Gamma} \lambda  - \left( \tfrac{1}{2}Id + K_{Y}^{\Gamma}  \right) u &= - \widetilde{u},  \\
 \left( \tfrac{1}{2} Id +  K_{Y}^{',\Gamma}\right)  \lambda + \left(W_{Y}^{\Gamma} + \tau_B Id\right)  u  &= \tau_B  \widetilde{u},
\end{align*}
where $K'$ and $W$ are the adjoint double-layer potentials and hypersingular operator for the Yukawa kernel.

Both, the interior and exterior problems have unique solutions for given $\widetilde{u}$. In order to find $\widetilde{u}$ we solve the following problem on $\Gamma$
$$
\tau_B \left( u_s - \widetilde{u}\right) + \tfrac{\tau_F}{h} \left( u_m - \widetilde{u}\right) - \left(\frac{\epsilon_m}{\epsilon_s} \partial_n u -\lambda\right)  = 0.
$$
\end{itemize}

### 2022-03-15 results.tex
Updated plots and manuscript
I attach plots with preconditioner and without. It give a little more understanding of behaviour of each method, at least I think so. We should discuss that.
In terms of manuscript, I am sorry, it is quite rough draft, but I am struggling with what should be included in terms of mathematics and derivation of methods.

The Kirkwood sphere [Kirkwood1934] is a standard benchmarck for the Poisson-Boltzmann equation in molecular electrostatics. 
Here, we considered a spherical cavity with radius $R=2$ \AA, with three charges ($q_1$=1, $q_2$=1, and $q_3$=0.75) placed at $\mathbf{x}_1=(1,0,0)$, $\mathbf{x}_2=(0.7,0.7,0)$, and $\mathbf{x}_3=(-0.5,-0.5,0)$.
Figure \ref{fig:error_sphere} shows the error convergence of the standard and hybrid FEM-BEM approaches, and a reference BEM implementation, to the analytical solution ($\Delta G_{solv}= -164.33752347454015$ kcal/mol). 
In these runs, the FEM mesh was generated using Bempp (check?) from a surface discretization with W, X, Y, and Z vertices per \AA$^2$ on the SES, which was also used for the BEM runs. 
For the hybrid FEM-BEM coupling approach, we used $\tau$=10.

@@

As a more realistic test, we assessed the performance of the standard and hybrid FEM-BEM coupling techniques against a purely BEM implementation for arginine [cite arg]. 
We generated surface meshes containing W, X, Y, and Z vertices per \AA$^2$ with Nanoshaper [Rocchia].
These meshes were inputs to our purely BEM solver, and to create the FEM mesh with pyGAMer [pygamer], invoking \texttt{tetgen} with the string \texttt{abcd}.
For the hybrid FEM-BEM coupling approach, we used $\tau$=10.

@@

\begin{figure}
\centering
   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_error.png}
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_iter.png}
  \includegraphics[width=0.6\linewidth]{Arginine_const_coeff_time.png}
\caption{Time-to-solution for arginine with a constant permittivity. %(1) fix title, (2) can we put them in a single plot, (3) 
maybe we could also add a "error" wrt extrapolation, or a plot with energy for each mesh?}
\label{fig:arg_contant_time}
\end{figure}

\begin{figure}
\centering
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_error.png}
  \includegraphics[width=0.6\linewidth]{Arginine_const_coeff_iter.png}
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_time.png}
\caption{Iteration count for arginine with a constant permittivity. %(1) fix title, (2) can we put them in a single plot, (3) hybrid is internal or external iterations? maybe we should present the total count?
}
\label{fig:arg_contant_iter}
\end{figure}

>>>>

The Kirkwood sphere [Kirkwood1934] is a standard benchmarck for the Poisson-Boltzmann equation in molecular electrostatics. 
Here, we considered a spherical cavity with radius $R=2$ \AA, with three charges ($q_1$=1, $q_2$=1, and $q_3$=0.75) placed at $\mathbf{x}_1=(1,0,0)$, $\mathbf{x}_2=(0.7,0.7,0)$, and $\mathbf{x}_3=(-0.5,-0.5,0)$.
Figure \ref{fig:error_sphere} shows the error convergence of the standard and hybrid FEM-BEM approaches, and a reference BEM implementation, to the analytical solution ($\Delta G_{solv}= -164.33752347454015$ kcal/mol). 
In these runs, the FEM mesh was generated using Bempp (check?) from a surface discretization with W, X, Y, and Z vertices per \AA$^2$ on the SES, which was also used for the BEM runs. 
For the hybrid FEM-BEM coupling approach, we used $\tau$=5.

@@

As a more realistic test, we assessed the performance of the standard and hybrid FEM-BEM coupling techniques against a purely BEM implementation for arginine [cite arg]. 
We generated surface meshes containing W, X, Y, and Z vertices per \AA$^2$ with Nanoshaper [Rocchia].
These meshes were inputs to our purely BEM solver, and to create the FEM mesh with pyGAMer [pygamer], invoking \texttt{tetgen} with the string \texttt{abcd}.
For the hybrid FEM-BEM coupling approach, we used $\tau$=5.

@@

\begin{figure}
\centering
   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_error.png}
   \includegraphics[width=0.45\linewidth]{No_prec_Arginine_const_coeff_error.png}
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_iter.png}
  \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_time.png}
  \includegraphics[width=0.45\linewidth]{No_prec_Arginine_const_coeff_time.png}
\caption{Time-to-solution for arginine with a constant permittivity (left: Preconditioned systems, right: No preconditioner).  %(1) fix title, (2) can we put them in a single plot, (3) 
%maybe we could also add a "error" wrt extrapolation, or a plot with energy for each mesh?
}
\label{fig:arg_contant_time}
\end{figure}

\begin{figure}
\centering
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_error.png}
  \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_iter.png}
  \includegraphics[width=0.45\linewidth]{No_prec_Arginine_const_coeff_iter.png}
%   \includegraphics[width=0.45\linewidth]{Arginine_const_coeff_time.png}
\caption{Iteration count for arginine with a constant permittivity (left: Preconditioned systems, right: No preconditioner). %(1) fix title, (2) can we put them in a single plot, (3) hybrid is internal or external iterations? maybe we should present the total count?
}
\label{fig:arg_contant_iter}
\end{figure}
